% 07/09/99
% $Id: ant32_tutorial.tex,v 1.7 2002/04/17 20:05:59 ellard Exp $

\documentclass[11pt,makeidx,psfig]{book}
\usepackage{ifthen}
\usepackage{makeidx}
\usepackage{alltt}

% \input{macros}
\input{dan}
\input{ant-macros}
\input{ant-spec}
\input{ant32-ver}

\newtheorem{alghead}{Algorithm}[chapter]
\newtheorem{codehead}{Program}[chapter]

\makeindex

\setlength{\textheight}{8.0in}
\setlength{\textwidth}{6.5in}
\setlength{\oddsidemargin}{0.0in}
\setlength{\evensidemargin}{0.0in}
\raggedbottom

\title{\Huge\bf Ant-32 \\ Assembly Language Tutorial \\
		(for version \input{../CurrVersion}) }

\author{Daniel J. Ellard, Penelope A. Ellard}

\begin{document}

\frontmatter
\maketitle

{\bf Copyright 2001-2002 by the President and Fellows of Harvard College. }

\tableofcontents

\chapter{Preface}

This document contains a brief tutorial for Ant-32 assembly
language programming, the assembly language utilities, and a
description of the general Ant-32 instruction set architecture. 
A complete specification of the Ant-32 architecture, including the
exception handling mechanisms, the MMU, and related issues, is given
in the {\em The Ant-32 Architecture (version \antVersion)},
which is provided as part of the Ant distribution and is also
available from the Ant web site ({\tt www.ant.harvard.edu}).

The Ant-32 architecture is a 32-bit RISC architecture designed
specifically for pedagogical purposes.  It is intended to be useful to
teaching a broad variety of topics, including machine architecture,
assembly language programming, compiler code generation, operating
systems, and VLSI circuit design and implementation.

\section{Outline}

\begin{itemize}

\item Chapter \ref{ant-asm-tut-intro} gives a tutorial for
	Ant-32 assembly language programming.  After reading this
	chapter, the reader should be able to write simple
	Ant-32 assembly language programs.

\item Chapter \ref{advanced-programming-chap} gives a tutorial for
	implementing function calls and related techniques in
	Ant-32 assembly language.

\item Chapter \ref{ant32-inst} gives a summary of the Ant-32
	instruction set.

\item Appendix \ref{rom-chapter} describes the default machine
	configuration, including the basic boot sequence and utility
	functions provided in the ROM.

\item Appendix \ref{ant32-asm-chapter} documents the assembler
	directives provided by the {\tt aa32} assembler.

\end{itemize}

\mainmatter

\input{tut32}

\chapter{Advanced Ant-32 Programming}
\label{advanced-programming-chap}
\input{reg32}

\input{t32-inst}

\appendix

\chapter{The Default Machine and ROM Routines}
\label{rom-chapter}
\input{rom32}

\chapter{Ant-32 Assembler Reference}
\label{ant32-asm-chapter}
\input{ant32-asm}

\backmatter

\printindex

\end{document}
