% 07/09/99
% $Id: ant8_tutorial.tex,v 1.13 2002/04/17 20:05:59 ellard Exp $

\documentclass[11pt,makeidx,psfig]{book}
\usepackage{ifthen}
\usepackage{makeidx}

% \input{macros}
\input{dan}
\input{ant-macros}

\newtheorem{alghead}{Algorithm}[chapter]
\newtheorem{codehead}{Program}[chapter]

\makeindex

\setlength{\textheight}{8.0in}
\setlength{\textwidth}{6.5in}
\setlength{\oddsidemargin}{0.0in}
\setlength{\evensidemargin}{0.0in}
\raggedbottom

% Define some variables that control which version of Ant-8 is being
% documented...

\title{\Huge\bf {\sc Ant-8} \input{../CurrVersion} \\ Assembly Language Tutorial }

\begin{document}

\frontmatter
\maketitle

{\bf Copyright 1996-2002 by the President and Fellows of Harvard College. }

\tableofcontents

\chapter{Preface}

This document is a brief tutorial for {\sc Ant-8} assembly language
programming, and a description of the {\sc Ant-8} architecture.

{\sc Ant-8} is a small and simple {\sc RISC} architecture that was
created as a teaching tool for students in introductory computer
programming and computer architecture courses.  It is simple enough so
that the entire architecture can be learned and understood in a few
hours by a novice programmer and small enough so that the entire state
of the machine can be displayed on a single 24-by-80 text screen.  The
{\sc Ant-8} architecture is also realistic enough, however, to give
students a real understanding of computer architecture and how their
programs are executed.

The first chapter gives a tutorial for {\sc Ant-8} assembly language
programming.  After reading this chapter, students should be able to
write simple {\sc Ant-8} assembly language programs.

The second chapter gives an overview specification of the {\sc Ant-8}
architecture.  After reading this chapter, students should be
able to create their own implementation of the {\sc Ant-8} (as a
software simulation, or as hardware).  More advanced students should
also be able to write {\sc Ant-8} debugging tools or even their own
assembler for {\sc Ant-8} assembly language after reading this chapter.

\mainmatter

\input{tutorial}

% \chapter{Functions}
% \input{func}

\input{tutex}

\input{ant-chap}

\backmatter

\printindex

\end{document}
